Management induced changes of soil organic carbon on global croplands

### 2020-07-14 SOCbudgetPaper/methods.Rmd

## Carbon Stocks following (new) Tier 2 method (50) {#sec:carbonbudget}

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{eq}$ are given by 
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with $C^{\textrm{in}}$ being carbon inputs to the soil, $k$ denoting the soil organic carbon decay rate; as well as $i$ representing grid cell indices and $t$ years. We use for our calculations the Tier 2 steady-state method, which assumes three soil carbon sub-pools (active, slow and passive) and interactions between them. Annual carbon inflow to each sub-pool (see \@ref(sec:carboninputs)) and annual decay rates (see \@ref(sec:tier2)) of each sub-pool are the key components to determining steady-state SOC stocks.

#### Carbon Inputs to the Soil {#sec:carboninputs}

We account for different carbon input sources depending on the two land-use types we distinguish, croplands and natural vegetation (see table \@ref(tab:datasourceinputs)). Carbon sources for cropland are recycled crop residues, below ground biomass of crops (for both see \@ref(sec:residues)) and recycled manure (see \ref(sec:livst_manure)); for natural vegetation litterfall (\ref(@LPJmL4_1)) litterfall is the only source of carbon.

@@

#### Soil Organic Carbon decay (300) {#sec:tier2}

@@

\begin{equation}
\begin{aligned}
& k_{active,t,i}  & = &~ k_{active}  ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ & \cdot~ sf_{t,i}\\
& k_{slow,t,i}    & = &~ k_{slow}    ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ &\\
& k_{passive,t,i} & = &~ k_{passive} ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ & ~ &
\label{eq:decayrates}
\end{aligned}
\end{equation}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall\quad\quad lut, sub, i.
\label{eq:steadystate}
\end{equation}

Reformulating equation \ref(eq:steadystate), we obtain a massbalance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation \ref(eq:inoutflow))}}  \qquad \forall\quad lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

## Agricultural management data on 0.5 degree resolution	(50) {#sec:agrimanagement}

@@

#### Crop Production
Using half-degree yield data from LPJmL ([@LPJmL4_1]) as well as half-degree cropland patterns (see \@ref(#sec:landuse)) we compile crop group specific half-degree production patterns. We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production ([@FAOSTAT]). Note that by using physical cropland areas we account for multiple crop harvest events as well as for fallows.

#### Crop Residue Production
Crop residue production and management is based on a revised methodology of ([@bodirsky2012]) and key aspects are explained here given its central role for soil carbon modelling. Starting from harvested crop production ($CP$) estimates and their respective harvested crop area ($CA$), we estimate above-ground ($AGR$) and below-ground ($BGR$) residual biomass using crop group ($cg$) specific harvest indices ($HI$) and root:shoot ratios ($RS$) as follows

\begin{equation}
\begin{aligned}
AGR & = CP \cdot HI_{\textrm{slope}} + CA \cdot HI_{\textrm{intercept}}\qquad & \textrm{and} \\
BGR & = (CP + AGR) \cdot RS \qquad                                            & \forall\quad cg, i, t.
\label{eq:resbiomass}
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the harvest index into a yield and an area dependent fraction ([@ipcc_2006_2006]). Note that, deviating from [@bodirsky2012], we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased amounts on fallow land. 
We assume that all BGR are recycled to the soil, whereas AGR can be burned or harvested for other purposes such as feeding animals ([@weindl]), fuel or for material use.

##### Burned Residues
A fixed share of the AGR is assumed to be burned on field depending on the per-capita income of the country. Following [@smil1999]) we assume 25% burn share for low-income countries according to worldbank definitions ($<\,1000\,\tfrac{USD}{yr\,cap}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr\,cap}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group 80–90% of the residue carbon burned on the fields are lost within the combustion process ([@ipcc_2006_2006]).

##### Residue Usage
We compile out of our 17 crop groups, three residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residue with no double use (see \@ref(append:Table_kcr2kres)). Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed baskets (see [@weindl]), taking available AGR biomass as well as livestock productivity into account. We estimate, for low-income countries,  a material-use share for straw residues of 5% and a fuel-share of 10% for all used residues groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumend, and middle-income countries use shares are linearly interpolated based on per-capita income. The remaining AGR as well as all BGR are assumend to be recycled to the soil. We limit high recycling shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

#### Dry Matter to Carbon Transformation
To transform dry matter estimates into carbon, we compiled crop-group and plant part specific carbon to dry matter (c:dm) ratios (see \@ref(append:Table_c2dm)) (@Ma2018).

@@

#### Livestock Distribution
To disaggregate country level FAOSTAT livestock production values to the half-degree scale, we use the following rule-based assumptions, drawing from by the approach of [@gilbert], and uses feed basket assumptions based on a revised methodology from @weindl. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
For ruminants, we assume that livestock is located where the production of feed occurs. We distingush between grazed pasture, which is converted into livestock products from extensive systems; and all other (crop-based) feeds, which we consider to be consumend in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to divide between intensive and extensive production systems. For low-income countries, we assume extensive production systems. We locate them according to the share of built-up areas based on the idea that these animals are held in subsistence or small-holder farming systems with a high labour per animal ratio. Intensive production associated with higher income countries, is distributed within a country using again share of crop-based feed production,  assuming that feed availability is the most driving factor for livestock location.

#### Manure Excretion, Storage and Recycling
Manure production and management is based on a revised methodology of ([@bodirsky2012]) and is presented here due to its central role in soil carbon modelling. Based on the gridded livestock distribution we calculate excretions by estimating the nitrogen balance of the livestock system on the basis of comprehensive livestock feed baskets ([@weindl]), assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios (given by [(@ipcc_2019_2019)]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure orginated from pasture feed intake is excreted directly to pastures and rangelands (pasture grazing), deducting manure collected as fuel. Manure fuel shares are estimated using IPCC default values (@ippc_2006_2006). 
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumend and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually recycled to the soil, we account for carbon losses during storage and recycling shares in different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not recycled. For manure stored in different animal waste management systems we compiled carbon loss rates partly depending on the nitrogen loss rates as specified in [@bodirsky2012] (see \@ref(append:Table_clossAWMS))). We limit high application shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

### Irrigation (100)

@@

### Tillage (100)

@@

\begin{equation}
SOC^{\text{target}}_{i} = \sum_{c,m} T_{c,i} \cdot SOC^{\text{ref}} \cdot F_{\text{LU}_{c,m}} \cdot F_{\text{MG}_{c,m}} \cdot F_{\text{I}_{c,m}} \cdot A_{m,i} \qquad\forall\quad t,
\label{eq:tier1}
\end{equation}

@@

<!-- KK: Representation of climate zones and grid cells is not consistent and has to be reformulated. -->

>>>>

## Carbon Stocks following (new) Tier 2 method {#sec:carbonbudget}

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{eq}$ are given by 
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being carbon inputs to the soil, $k$ denoting the soil organic carbon decay rate; as well as $i$ representing grid cell indices and $t$ years. We use for our calculations the Tier 2 steady-state method, which assumes three soil carbon sub-pools (active, slow and passive) and interactions between them. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are the key components to determining steady-state SOC stocks.

#### Carbon Inputs to the Soil {-}

We account for different carbon input sources depending on the two land-use types we distinguish, croplands and natural vegetation (see table \@ref(tab:datasourceinputs)). Carbon sources for cropland are recycled crop residues, below ground biomass of crops (for both see \@ref(sec:residues)) and recycled manure (see \@ref(sec:livst_manure)); for natural vegetation litterfall ([@LPJmL4_1]) litterfall is the only source of carbon.

@@

#### Soil Organic Carbon decay {-}

@@

\begin{equation}
\begin{aligned}
& k_{active,t,i}  & = &~ k_{active}  ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ & \cdot~ sf_{t,i}\\
& k_{slow,t,i}    & = &~ k_{slow}    ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ &\\
& k_{passive,t,i} & = &~ k_{passive} ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ & ~ &
(\#eq:decayrates)
\end{aligned}
\end{equation}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall\quad\quad lut, sub, i.
(\#eq:SOCstate)
\end{equation}

Reformulating equation \@ref(eq:steadystate), we obtain a massbalance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation \@ref(eq:inoutflow))}}  \qquad \forall\quad lut, sub, i.
(\#eq:steadystate2budget)
\end{equation}

@@

## Agricultural management data on 0.5 degree resolution {#sec:agrimanagement}

@@

#### Crop Production {-}
Using half-degree yield data from LPJmL ([@LPJmL4_1]) as well as half-degree cropland patterns (see \@ref(#sec:landuse)) we compile crop group specific half-degree production patterns. We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production ([@FAOSTAT]). Note that by using physical cropland areas we account for multiple crop harvest events as well as for fallows.

#### Crop Residue Production {-}
Crop residue production and management is based on a revised methodology of ([@bodirsky2012]) and key aspects are explained here given its central role for soil carbon modelling. Starting from harvested crop production ($CP$) estimates and their respective harvested crop area ($CA$), we estimate above-ground ($AGR$) and below-ground ($BGR$) residual biomass using crop group ($cg$) specific harvest indices ($HI$) and root:shoot ratios ($RS$) as follows

\begin{equation}
\begin{aligned}
AGR & = CP \cdot HI_{\textrm{slope}} + CA \cdot HI_{\textrm{intercept}}\qquad & \textrm{and} \\
BGR & = (CP + AGR) \cdot RS \qquad                                            & \forall\quad cg, i, t.
(\#eq:resbiomass)
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the harvest index into a yield and an area dependend fraction ([@ipcc_2006_2006]). Note that deviating from [@bodirsky2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased amounts on fallow land. 
We assume that all BGR are recycled to the soil, whereas AGR can be burned or harvested for other purposes such as feeding animals ([@weindl]), fuel or for material use.

##### Burned Residues {-}
A fixed share of the AGR is assumed to be burned on field depending on the per-capita income of the country. Following [@smil1999]) we assume 25% burn share for low-income countries according to worldbank definitions ($<\,1000\,\tfrac{USD}{yr\,cap}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr\,cap}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group 80–90% of the residue carbon burned on the fields are lost within the combustion process ([@ipcc_2006_2006]).

##### Residue Usage {-}
We compile out of our 17 crop groups, three residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residue with no double use (see \@ref(append:Table_kcr2kres)). Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed baskets (see [@weindl]), taking available AGR biomass as well as livestock productivity into account. We estimate, for low-income countries,  a material-use share for straw residues of 5% and a fuel-share of 10% for all used residues groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumend, and middle-income countries use shares are linearly interpolated based on per-capita income. The remaining AGR as well as all BGR are assumend to be recycled to the soil. We limit high recycling shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

#### Dry Matter to Carbon Transformation {-}
To transform dry matter estimates into carbon, we compiled crop-group and plant part specific carbon to dry matter (c:dm) ratios (see \@ref(append:Table_c2dm)) (@Ma2018).

@@

#### Livestock Distribution {-}
To disaggregate country level FAOSTAT livestock production values to the half-degree scale, we use the following rule-based assumptions, drawing from by the approach of [@gilbert], and uses feed basket assumptions based on a revised methodology from @weindl. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
For ruminants, we assume that livestock is located where the production of feed occurs. We distingush between grazed pasture, which is converted into livestock products from extensive systems; and all other (crop-based) feeds, which we consider to be consumend in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to divide between intensive and extensive production systems. For low-income countries, we assume extensive production systems. We locate them according to the share of built-up areas based on the idea that these animals are held in subsistence or small-holder farming systems with a high labour per animal ratio. Intensive production associated with higher income countries, is distributed within a country using again share of crop-based feed production,  assuming that feed availability is the most driving factor for livestock location.

#### Manure Excretion, Storage and Recycling {-}
Manure production and management is based on a revised methodology of ([@bodirsky2012]) and is presented here due to its central role in soil carbon modelling. Based on the gridded livestock distribution we calculate excretions by estimating the nitrogen balance of the livestock system on the basis of comprehensive livestock feed baskets ([@weindl]), assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios (given by [(@ipcc_2019_2019)]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure orginated from pasture feed intake is excreted directly to pastures and rangelands (pasture grazing), deducting manure collected as fuel. Manure fuel shares are estimated using IPCC default values (@ippc_2006_2006). 
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumend and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually recycled to the soil, we account for carbon losses during storage and recycling shares in different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not recycled. For manure stored in different animal waste management systems we compiled carbon loss rates partly depending on the nitrogen loss rates as specified in [@bodirsky2012] (see \@ref(append:Table_clossAWMS))). We limit high application shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

### Irrigation

@@

### Tillage

@@

\begin{equation}
SOC^{\text{target}}_{i} = \sum_{c,m} T_{c,i} \cdot SOC^{\text{ref}} \cdot F_{\text{LU}_{c,m}} \cdot F_{\text{MG}_{c,m}} \cdot F_{\text{I}_{c,m}} \cdot A_{m,i} \qquad\forall\quad t,
(\#eq:tier1)
\end{equation}

@@

According to [@ipcc_2019_2019] the formulation of absolute changes like
<!-- plugin absolute chnage equation of refinement cropland chapter -->
might be neccessary as well.
